Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-10-01 analysis/paper/paper.tex
segudno borrador

\authors{Elio Campitelli
\correspondingauthor{Elio Campitelli,elio.campitelli@cima.fcen.uba.ar}
and Leandro Díaz
}

@@

Most authors describe the SAM as a zonally symmetric pattern, a fact
that is reflected not only in its name, but also in the various methods
used to characterise it. Of the several different indices presented in
the literature, many of them are based on zonal means of sea level
pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
the SAM index as the zonal mean sea level pressure differene between
40\degree S and 60\degree S, which is also the definition used by the
station-based index in \citet{marshall2003}. \citet{baldwin2009}
proposed defining the Northern and Annular modes as the leading EOF of
the zonally averaged geopotential height at each level.

Even though these indices are based on zonal averages, the spatial the
spatial structure of the SAM computed from them contains noticeable
deviations from zonal symmetry, particularly in the Pacific Ocean
region. These zonal asymmetries have not been widely studied, but
previous work suggest that they strongly modulate the regional impacts
of the SAM, going as far as reversing its relationship between
precipitation in South America \citep{silvestri2009}. At the very least,
the fact that the SAM is not entirely zonally symmetric hinders our
ability to reconstruct its historical variability prior to the
availabilty of dense observations in the Southern Hemisphere
\citep{jones2009}.

At least some of the variabilty associated with the zonal asymmetries of
the SAM is probably forced by the tropics. In particular, ENSO-like
variability affects the Southern Hemisphere extratopics through the
Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
whose wave train proyects strongly onto the zonal anomalies
corresponding to the SAM in the Pacific sector. And although the
relationship between ENSO and SAM is far from simple, tropical
influences on the SAM have been observed
\citep{fan2007, fogt2011, clem2013}.

Positive trends in SAM index have been documented by various researchers
using different indices mostly on boreal Summer and Autumn \citep[ and
references therein]{fogt2020}. These trends are thought of driven
primarily by stratospheric ozone depletion and understood in the context
of zonal mean variables. However, it's not clear how or if the
asymmetric component responds to this forcing, or whether its variabilty
could be masking influencing the observed trends independendly.

Similarly unclear are the specific impacts of the zonally asymmetric
component of the SAM. Positive phase of the SAM is associated with
generally colder temperatures over Antarctica and warmer temperatures at
higher latitudes \citep{jones2019} (and viceversa for negative SAM), but
there are significant deviations from this zonal mean response, notably
in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}. The
SAM signal in precipitation behaves similarly, although with even
greater deviation from zonal symmetry \citep{lim2016}. The importance of
zonal asymmetries of the SAM in these impacts have been studied in
certain regions. For example, the SAM-precipitation relationship in
Southeastern South America and Southern Brazil can be explained by the
PSA-like zonally asymetric circulation associated with the SAM
\citep{silvestry2009, rosso2018}. \citet{fan2007} also found that
precipitation in East Asia was impacted by the variabilty of only the
Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal
variability of the asymmetric component of the SAM with the exception of
\citet{fogt2012}. However, their methods based on composites of positive
and negative SAM events leads to some issues, such as spatial patterns
derived from as little as 4 cases and from imbalanced periods (for
example, 5 of the 7 cases in their DJF SAM+ composite are from later
than 1988, whereas all of the 8 years in their DJF SAM- composite are
from earlier than 1988). This is particularly important due to the
inhomogeneities in renalysis products prior to the satellite era and the
possible change in the asymmetric structure of the SAM
\citep{silvestri2009}.

@@

We describe the relationship between the SAM indices and temperature and
precipitation. We use temperature data from NOAA's Merged Land Ocean
Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
which blends land and ocean temperature analysis into a monthly global
grid 5\degree~ longitude by 5\degree~ latitude For precipitation, we use
monthly, 0.5\degree~ longitude by 0.5\degree~ latitude data from the
Global Precipitation Climatology Centre \citep{schneider2015}.

@@

Traditionally the Souther Annular Mode (SAM) is defined as de leading
Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential
height at lower levels \citep{ho2012}. Following \citet{baldwin2001}, we
extend that definition vertically and use the term SAM to refer to the
the leading EOF of the monthly anomalies of geopotential field south of
20\degree S at each level. We performed EOFs by computing the Singular
Value Decomposition of the data matrix consisting in 481 rows and 4176
columns (144 points of longitude and 29 points of latitude). We weighted
the values by the square root of the cosine of latitude to account for
the non-equal area of each gridpoint \citep{chung1999}.

To separate between the zonally symmetric and asymmetric components of
the SAM, we computed the zonal mean and anomalies of the full SAM
spatial pattern, as shown in Figure~\ref{fig:method} for 700hPa. The
full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of
the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and
symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM''
indices as the regression coefficients of the regression ofeach monthly
geopotential field on the respective patterns (weighting by the cosine
of latitude). The three indices are normalised by dividing them by the
standard deviation of the ``Full'' index at each level. As a result, the
magnitude between indices is comparable. However, only ``Full'' index
will have unit standard deviation per definition. From the regression,
we also use the explained variance of each pattern as a indication of
the degere of symmetry or asymmetry of each monthly field.

\begin{figure*}
\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height]{Spatial patterns of the first EOF of 700 hPa geopotential height. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.}\label{fig:method}
\end{figure*}

Our method assumes linearity in the asymmetric component of the SAM.
That is, we assume that zonal symmetries associated with positive SAM
are oposite and equal to the ones associated with negatie SAM.
\citet{fogt2012}'s composites (their Figure 4) suggest that this might
not be entirely valid, although we argue that much of that apparent
non-linearity is due to the heterogenous nature of the selected years
for constructing the composites. Using our data (from 1979 to 2018),
seasonal composites of zonal anomalies of 700 hPa geopotential height
for SAM+ (Full SAM index greater than 1 standard deviation) and SAM-
(smaller than negative 1 standard deviation) show relatively high
pattern correlations all seasons and are visually very linear (Figure
A9). Therefore, we belive that our method is at the very least a
reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are
assuming that the zonal anomalies of the SAM are the same in all
seasons. Geopotential zonal anomalies computed by projecting the first
EOF \emph{of each season} are very similar to each other (Figure A10)
and show pattern correlations between 0.65 (DJF with JJA) and 0.9
(between MAM and SON). Based on this, we believe that our initial
assumption is not unreasonable.

@@

We perform linear regression to quantify the association between the SAM
indices and other variables. Since the Asymmetric and Symmetric SAM
indices are significantly correlated with each other, to capture the
variability explained uniquely by each index we use one multiple linear
regression instead of two simple linear regressions. To obtain the
linear coefficients of a variable \(X\) (geopotential, temperature,
precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and
Symmetric SAM (\(SAM_s\)) we fit the ecuation

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
the time, \(\alpha\) and \(\beta\) are the linear coefficients, \(X_0\)
and \(\epsilon\) are the constant and error terms. From this equarion,
\(\alpha\) represents the (linear) association of \(X\) with the
variability of the Asymmetric SAM that is not explained by the
variability of the Symmetric SAM; in other words, it is proportional to
the partial correlation of \(X\) and the Asymmetric SAM, controlling for
the effect of the Symmetric SAM and viceversa for \(\beta\).

At 2.5\degree by 2.5\degree resolution, a single regression field is
composed of thousands of regressions. In such case, using naive p-values
to test for significance leads to misleading results
\citep{walker1914, katz1991}. While there are multiple proposed
solutions in the literature, \citet{wilks2016} suggests that adjunting
p-values by controlling for the False Discovery Rate
\citep{benjamini1995} is a simple and effective method to ameliorate
this issue. Therefore, p-values showed in regression fields are all
adjusted following \citet{benjamini1995}.

When performing a separate regression for each trimester (DJF, MAM, JJA,
SON) we first average the relevant variables to obtain a single value
for each year and each trimester.

@@

\begin{figure*}
\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa. To the left, probability density estimate of each index. Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
\end{figure*}

The temporal evolution of the Assymmetric and Symmetric SAM was firstly
asssesed. Figure \ref{fig:asymsam-timeseries} shows the corresponding
time series for 700 hPa and 50 hPa and their corresponding density
estimates. We selected these two levels as representative of the
tropospheric and stratospheric variability respectively. As will be
shown later, both indices are highly coherent within each atmospheric
layer, therefore is reasonable to take one level as representative of
each layer.

Month-to-month variability is evident for both indices, with noisy
variations in the low frequency. At first glance the series can be
distinguished by their distributions. Compared to the stratospheric
indices, the stratospheric indices are much more long-tailed; that is,
extreme values (both negative and positive) abound. The Asymmetric
series have both more variability in the higher frequencies than the
Symmetric series.

The stratospheric Symmetric SAM varies strongly with a two-year period,
which can be seen by spectral analysis (Figure A3). This might suggests
a link between stratospheric SAM variability and QBO. There is a local
peak at 2 years in the periodigram of the tropospheric Symmetric SAM
also, although it's not statistically significant. In the troposphere
the most significant peak of variability is found in the Asymmetric
index at around 3.6 months.

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric
and Symmetric time series appear to be correlated. Moreover, looking at
the extremes in the stratosphere, the Symmetric serie appears to lag the
Asymmetric series (see, for example, the positive events on late 1987).
We show these correlations, across all the levels of the reanalysis for
zero and -1 lag (Asymmetric index leading the Symmetric index), in
Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric
and Symmetric series are relatively constant throught the troposphere,
fluctiating between 0.39 and 0.45. One-month-lag correlations are
similarly constant but significantly reduced to around 0.17. In the
stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20 hPa
and then it increases again monotonically with height up to the
uppermost level of the reanalysis (although results near the top of the
models are to be interpreted with care). At the same time, one-month-lag
correlations increase with height. As a consequence, statospheric
Symmetric index tend to precede corresponding Asymmetric index.

\begin{figure}
\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric)]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).}\label{fig:cor-lev}
\end{figure}

\begin{figure*}
\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.}\label{fig:cross-correlation}
\end{figure*}

Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation
across levels for the Full, Symmetric and Asymmetric SAM indices. For
the Full SAM (panel a), high values below 100 hPa reflect the vertical
(zero-lag) coherency throughout the troposfere. Above 100 hPa
correlation between levels falls off more rapidly, indicating less
coherent (zero-lag) variability. Therefore, there is a non negligible
correlation between the troposphere and the lower-to-middle
stratosphere. Examining panels b and c, we see that the Asymemtric and
Symmetric SAM share the same high level of coherency in the troposphere
but they differ in their stratospheric behaviour. Stratospheric
coherency is stronger for the Asymmetric SAM than the Symemtric SAM. The
stratospheric Symmetric SAM seems to connect more strongly to the
trosposphere than the Asymmetric SAM.

\begin{figure*}
\includegraphics{trends-1} \caption[Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\% confidence interval.}\label{fig:trends}
\end{figure*}

The trends for each of the indices (Full, Symmetric, Assymetric) were
evaluate for the whole period 1979-2018 at each level (Figure
\ref{fig:trends}) for the whole year and separed by trimesters. The Full
SAM index presents a statistically significant trend (panel a.1) that
extends throught the troposphere up to about 50 hPa and reaches its
maximum value at 100 hPa. The seasonal trends (rest of column a)
indicate that positive trends are present in Autumn and particularly in
in Summer, where the 100 hPa maximum is much more defined. Positive
trends have been documented by previous studies \citep[e.g.][ and
references therein]{fogt2020} using indices of the SAM based on surface
or near-surface circulation.

By separating the SAM signal in its Asymmetric and Symmetric parts, we
can not only see that these trends are almost entirely due to the
Symmetric component (colum b vs.~column c), but in some cases the trends
become more clear. In Summer, the Asymmetric SAM has a statistically non
significant negative trend in the middle troposphere that obscures the
trend in the Full index; as a result, trends computed using only the
Symmetric component are more clear (compare the shading region in panel
a.2 and c.2). In Autumn, using the Symmetric SAM reveals a statistically
significant positive trend in the stratosphere that is not significant
using the Full index.

We stress that these are only linear trends during the whole period and
the absence of a statistically significant signal should not be taken as
evidence of no sistematic change. In particular, going back to Figure
\ref{fig:asymsam-timeseries}, we can see an evident change in the
stratospheric Asymemtric component (red line in panel a) between the
90's, when we see a dominance of extreme negative values, and the 00's,
when we see the inverse. This change is restricted to the Winter months:
the linear trend for JJA starting in 1990 for the Asymmetric component
at 50hPa is \(0.37 \pm 0.22\).

Figure \ref{fig:r-squared-trend} shows decadal trends for the explained
variance of each index. There is no evidencie of a significant trend in
the stratosphere. In the troposphere, there is a positive trend for the
Asymmetric SAM and no significant trend for the Symmetric SAM. This
suggest that the SAM has become more asymmetric in the period from 1979
to 2018. The change is slight, though; of the order of 1\% icreased
explained variance per decade.

@@

\begin{figure*}
\includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM]{Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index. }\label{fig:2d-regr}
\end{figure*}

To show if, and to what extent, the Asymmetric and Symmetric SAM
inidices indeed capture the asymmetric and symmetric compoment of the
SAM respectively, we computed the spatial regression of geopotential
height anomalies on these indices and the Full SAM index. Figure
\ref{fig:2d-regr} shows these regressions. Regression coefficients in
column a are computed using the Full SAM. Regression coefficients in
columns b and c are computed using multiple regression using the
Asymmetric and Symmetric indices at the same time. Thus, they are to be
interpreted as the patterns associated with each index, removing the
variability (linearly) explained by the other index.

In the stratosphere, the spatial pattern associated with the Full SAM is
more clearly dominated by a zonally symmetric, monopolar structure
(panel a.1) which is, however, not perfectly centered in the South Pole.
The monopole obtained by multiple regression with the Asymmetric and
Symmetric SAM (panel c.1) is much more symmetric and the shift from
total symmetry is captured by the regression pattern of the Asymmetric
SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the
Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel
b.1).

>>>>

\authors{Elio Campitelli
\correspondingauthor{Elio Campitelli,elio.campitelli@cima.fcen.uba.ar}
and Leandro B. Díaz
}

@@

Most authors describe the SAM as a zonally symmetric pattern, a fact
that is reflected not only in its name, but also in the various methods
used to characterise it. Of the several different indices presented in
the literature, many of them are based on zonal means of sea level
pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
the SAM index as the zonal mean sea level pressure difference between
40\degree S and 60\degree S, which is also the definition used by the
station-based index in \citet{marshall2003}. \citet{baldwin2009}
proposed defining the Northern and Annular modes as the leading EOF of
the zonally averaged geopotential height at each level.

Even though these indices are based on zonal averages, the spatial the
spatial structure of the SAM computed from them contains noticeable
deviations from zonal symmetry, particularly in the Pacific Ocean
region. These zonal asymmetries have not been widely studied, but
previous work suggest that they strongly modulate the regional impacts
of the SAM, going as far as reversing its relationship between
precipitation in South America \citep{silvestri2009}. At the very least,
the fact that the SAM is not entirely zonally symmetric hinders our
ability to reconstruct its historical variability prior to the
availability of dense observations in the Southern Hemisphere
\citep{jones2009}.

At least some of the variability associated with the zonal asymmetries
of the SAM is probably forced by the tropics. In particular, ENSO-like
variability affects the Southern Hemisphere extratopics through the
Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
whose wave train protects strongly onto the zonal anomalies
corresponding to the SAM in the Pacific sector. And although the
relationship between ENSO and SAM is far from simple, tropical
influences on the SAM have been observed
\citep{fan2007, fogt2011, clem2013}. In particular, \citet{fan2007}
computed SAM indices of the Western and the Eastern Hemisphere
separately and found that they were much more correlated if the (linear)
signal of the ENSO was removed. While this relates to temporally
coherent variability of the two hemisphere and not necessarily to zonal
asymmetries in the associated spatial patterns, it is nonetheless
consistent of the ENSO playing a crucial role in zonal asymmetries of
the SAM.

Positive trends in SAM index have been documented by various researchers
using different indices mostly on boreal Summer and Autumn \citep[ and
references therein]{fogt2020}. These trends are thought of driven
primarily by stratospheric ozone depletion and understood in the context
of zonal mean variables. However, it's not clear how or if the
asymmetric component responds to this forcing, or whether its
variability could be masking influencing the observed trends
independently.

Similarly unclear are the specific impacts of the zonally asymmetric
component of the SAM. Positive phase of the SAM is associated with
generally colder temperatures over Antarctica and warmer temperatures at
higher latitudes \citep{jones2019} (and vice versa for negative SAM),
but there are significant deviations from this zonal mean response,
notably in the Antarctic Peninsula and the South Atlantic
\citep{fogt2012}. The SAM signal in precipitation behaves similarly,
although with even greater deviation from zonal symmetry
\citep{lim2016}. The importance of zonal asymmetries of the SAM in these
impacts have been studied in certain regions. For example, the
SAM-precipitation relationship in Southeastern South America and
Southern Brazil can be explained by the PSA-like zonally asymmetric
circulation associated with the SAM \citep{silvestri2009, rosso2018}.
\citet{fan2007} also found that precipitation in East Asia was impacted
by the variability of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal
variability of the asymmetric component of the SAM with the exception of
\citet{fogt2012}. However, their methods based on composites of positive
and negative SAM events leads to some issues, such as spatial patterns
derived from as little as 4 cases and from imbalanced periods (for
example, 5 of the 7 cases in their DJF SAM+ composite are from later
than 1988, whereas all of the 8 years in their DJF SAM- composite are
from earlier than 1988). This is particularly important due to the
inhomogeneities in reanalysis products prior to the satellite era and
the possible change in the asymmetric structure of the SAM
\citep{silvestri2009}.

@@

To describe the relationship between the SAM indices and temperature and
precipitation, we use temperature data from NOAA's Merged Land Ocean
Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
which blends land and ocean temperature analysis into a monthly global
grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall
0.5\degree~longitude by 0.5\degree~latitude data from the Global
Precipitation Climatology Centre \citep{schneider2015}.

@@

Traditionally the Souther Annular Mode (SAM) is defined as de leading
Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential
height at lower levels \citep{ho2012}. Following \citet{baldwin2001}, we
extend that definition vertically and use the term SAM to refer to the
the leading EOF of the monthly anomalies of geopotential field south of
20\degree S at each level. We performed EOFs by computing the Singular
Value Decomposition of the data matrix consisting in 481 rows and 4176
columns (144 points of longitude and 29 points of latitude). We weighted
the values by the square root of the cosine of latitude to account for
the non-equal area of each gridpoint \citep{chung1999}.

To separate the zonally symmetric and asymmetric components of the SAM,
we computed the zonal mean and anomalies of the full SAM spatial
pattern, as shown in Figure \ref{fig:method} for 700 hPa. The full
spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the
zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric
(\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then compute the
``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the
regression coefficients of the regression of each monthly geopotential
field on the respective patterns (weighting by the cosine of latitude).
The three indices are normalized by dividing them by the standard
deviation of the ``Full'' index at each level. As a result, the
magnitudes between indices are comparable. However, only ``Full'' index
will have unit standard deviation per definition. From the regression,
we also use the explained variance of each pattern as an indicator of
the degree of zonally symmetry or asymmetry of each monthly field.

\begin{figure*}
\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.}\label{fig:method}
\end{figure*}

Our method assumes linearity in the asymmetric component of the SAM.
That is, we assume that zonal symmetries associated with positive SAM
are almost opposite and of the same magnitude to the ones associated
with negative SAM. \citet{fogt2012}'s composites (their Figure 4)
suggest that this might not be entirely valid, although we argue that
much of that apparent non-linearity is due to the heterogenous nature of
the selected years for constructing the composites. Using our data (from
1979 to 2018), seasonal composites of zonal anomalies of 700 hPa
geopotential height for SAM+ (Full SAM index greater than 1 standard
deviation) and SAM- (less than minus 1 standard deviation) show
relatively high pattern correlations for all seasons and are visually
very linear (Figure A9). Therefore, we believe that our method is a
reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are
assuming that the zonal anomalies of the SAM are the same in all seasons
--December to February (DJF), March to May (MAM), June to August (JJA)
and September to November (SON). Geopotential zonal anomalies computed
by projecting the first EOF \emph{of each season} are very similar to
each other (Figure A10) and show pattern correlations between 0.65 (DJF
with JJA) and 0.9 (between MAM and SON). Based on this, we believe that
our initial assumption is not unreasonable.

@@

We perform linear regressions to quantify the association between the
SAM indices and other variables. Since the Asymmetric and Symmetric SAM
indices are significantly correlated with each other, to capture the
variability explained uniquely by each index we use one multiple linear
regression instead of two simple linear regressions. To obtain the
linear coefficients of a variable \(X\) (geopotential, temperature,
precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and
Symmetric SAM (\(SAM_s\)) we fit the equation

@@

where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
the time, \(\alpha\) and \(\beta\) are the linear regression
coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
From this equation, \(\alpha\) represents the (linear) association of
\(X\) with the variability of the Asymmetric SAM that is not explained
by the variability of the Symmetric SAM; i.e.~it is proportional to the
partial correlation of \(X\) and the Asymmetric SAM, controlling for the
effect of the Symmetric SAM and vice versa for \(\beta\). When
performing a separate regression for each trimester (DJF, MAM, JJA,
SON), we average of the relevant variables seasonally for each year and
trimester before computing the regression.

At 2.5\degree by 2.5\degree resolution, a single regression field is
composed of thousands of regressions. In such case, using p-values to
test for significance leads to misleading results
\citep{walker1914, katz1991}. While there are multiple proposed
solutions in the literature, \citet{wilks2016} suggests that adjusting
p-values by controlling for the False Discovery Rate
\citep{benjamini1995} is a simple and effective method to ameliorate
this issue. Therefore, p-values showed in regression fields are all
adjusted following \citet{benjamini1995}.

We computed linear trends by Ordinary Least Squares and the 95\%
confidence interval assuming a t-distribution of the appropriate
residual degrees of freedom.

To compute the amplitude of the zonal waves we computed the Fourier
transform of the spatial field at each latitude circle.

\subsubsection{Computation procedures}

We performed all analysis in this paper using the R programming language
\citep{rcoreteam2020}, using the data.table package \citep{dowle2020}
and the metR package \citep{campitelli2020}. All graphics are made using
ggplot2 \citep{wickham2009}. We downloaded data from ranalysis using the
ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi
package \citep{albers2020}. The paper was rendered using knitr and
rmarkdown \citep{xie2015, allaire2019}.

A version-controlled repository of the code used to create this analysis
can be found at https://github.com/eliocamp/asymsam.

@@

\begin{figure*}
\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa. To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \cite{sheather1991}. Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
\end{figure*}

We first asses the temporal evolution of the Asymmetric SAM and
Symmetric SAM. Figure \ref{fig:asymsam-timeseries} shows the
corresponding time series for 700 hPa and 50 hPa and their corresponding
density estimates. We selected these two levels as representative of the
tropospheric and stratospheric variability respectively. As will be
shown later, both indices are highly coherent within each atmospheric
layer, therefore is reasonable to take one level as representative of
each layer.

Month-to-month variability is evident for both indices, with noisy
variations in the low frequency. At first glance the series can be
distinguished by their distributions. Compared to the tropospheric
indices, the stratospheric indices are much more long-tailed; that is,
extreme values (both negative and positive) abound. The Asymmetric SAM
series have both more variability in the higher frequencies than the
Symmetric SAM series.

The stratospheric Symmetric SAM varies strongly with a two-year period,
which can be seen by spectral analysis (Figure A3). This might suggests
a link between stratospheric SAM variability and the Quasi-Biennial
Oscillation \citep{baldwin2001b}. There is a local peak at 2 years in
the periodogram of the tropospheric Symmetric SAM also, although it's
not statistically significant. In the troposphere the most significant
peak of variability is found in the Asymmetric index at around 3.6
months.

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric
SAM and Symmetric SAM time series appear to be correlated. Moreover,
looking at the extremes in the stratosphere, the Symmetric SAM series
appears to lag the Asymmetric SAM series (see, for example, the positive
events on late 1987). We show these correlations, across all the levels
of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the
Symmetric SAM index), in Figure \ref{fig:cor-lev}. Zero-lag correlations
between the Asymmetric SAM and Symmetric SAM series are relatively
constant throughout the troposphere, fluctuating between 0.39 and 0.45.
One-month-lag correlations are similarly constant but significantly
reduced to around 0.17. In the stratosphere, zero-lag correlations drop
to a minimum of 0.21 at 20 hPa and then it increases again monotonically
with height up to the uppermost level of the reanalysis (although
results near the top of the models are to be interpreted with care). At
the same time, one-month-lag correlations increase with height. As a
consequence, stratospheric Asymmetric SAM index tend to precede
corresponding Symmetric SAM index.

\begin{figure}
\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.}\label{fig:cor-lev}
\end{figure}

\begin{figure*}
\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.}\label{fig:cross-correlation}
\end{figure*}

Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation
across levels for the Full, Symmetric and Asymmetric SAM indices. For
the Full SAM (panel a), high values below 100 hPa reflect the vertical
(zero-lag) coherency throughout the troposphere. Above 100 hPa
correlation between levels falls off more rapidly, indicating less
coherent (zero-lag) variability. Therefore, there is a non negligible
correlation between the troposphere and the lower-to-middle
stratosphere. Examining panels b and c, we see that the Asymmetric and
Symmetric SAM share the same high level of coherency in the troposphere
but they differ in their stratospheric behaviour. Stratospheric
coherency is stronger for the Asymmetric SAM than the Symmetric SAM. The
stratospheric Symmetric SAM seems to connect more strongly to the
troposphere than the Asymmetric SAM.

\begin{figure*}
\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
\end{figure*}

The linear trends for each of the indices (Full SAM, Symmetric SAM and
Asymmetric SAM) were evaluated for the complete period 1979-2018 at each
level (Figure \ref{fig:trends}) for the whole year and separated by
trimesters. The Full SAM index presents a statistically significant
trend (panel a.1) that extends throughout the troposphere up to about 50
hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest
of column a) indicate that positive trends are present in Autumn and
particularly in Summer, where the 100 hPa maximum is much more defined.
In Winter and Spring, we detect no statistically significant trend. This
is consistent with the results of previous studies, which find clear
positive trends in Summer, weaker in Autumn and no trends in the other
seasons \citep[e.g.][ and references therein]{fogt2020} using indices of
the SAM based on surface or near-surface circulation.

By separating the SAM signal in its asymmetric and symmetric parts, we
can not only see that these trends are almost entirely due to the
symmetric component (column b vs.~column c), but in some cases the
trends become more clear. In Summer, the Asymmetric SAM has a
statistically non significant negative trend in the middle troposphere
that obscures the trend in the Full SAM index; as a result, trends
computed using only the Symmetric component are more clear (compare the
shading region in panel a.2 and c.2). In Autumn, the Symmetric SAM
reveals a statistically significant positive trend in the stratosphere
that is not significant using the Full SAM index.

We stress that these are only linear trends during the whole period and
the absence of a statistically significant signal should not be taken as
evidence of no systematic change. In particular, going back to Figure
\ref{fig:asymsam-timeseries}, we can see an evident change in the
stratospheric Asymmetric component (red line in panel a) between the
90's, when we see a dominance of extreme negative values, and the 00's,
when we see the inverse. This change is restricted to the Winter months:
the linear trend for Winter starting in 1990 for the Asymmetric
component at 50hPa is \(0.37 \pm 0.22\).

Figure \ref{fig:r-squared-trend} shows decadal trends for the explained
variance of each index. There is no evidence of a significant trend in
the stratosphere. In the troposphere, there is a positive trend for the
Asymmetric SAM and not significant trend for the Symmetric SAM. This
suggest that the SAM has become more asymmetric in the period from 1979
to 2018. However, the change is slight, around 1\% increased explained
variance per decade.

@@

\begin{figure*}
\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices. Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index. }\label{fig:2d-regr}
\end{figure*}

To show if, and to what extent, the Asymmetric and Symmetric SAM indices
indeed capture the asymmetric and symmetric component of the SAM
respectively, we computed the spatial regression of geopotential height
anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa
levels. Figure \ref{fig:2d-regr} shows these regressions. Regression
coefficients in column a are computed using the Full SAM. Regression
coefficients in columns b and c are computed using multiple regression
using the Asymmetric and Symmetric indices at the same time. Thus, they
are to be interpreted as the patterns associated with each index,
removing the variability (linearly) explained by the other index.

In the stratosphere, the spatial pattern associated with the Full SAM is
more clearly dominated by a zonally symmetric, monopolar structure
(panel a.1) which is, however, not perfectly centred in the South Pole.
The monopole obtained by the regression pattern for Symmetric SAM (panel
c.1) is much more symmetric and the shift from total symmetry is
captured by the regression pattern of the Asymmetric SAM as a wave-1
with maximum anomalies above the Belinghausen Sea on the Western
Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
